\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-16 paper.tex
save

\section{Problem statement}

We consider a family of parameterized densities $p_\mathbf{\theta}(\mathbf{x})$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\mathbf{\theta}$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p_\mathbf{\theta}(\mathbf{x}) = \int p_\mathbf{\theta}(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in \mathbb{R}^m$ is a latent variable providing an external source
of randomness.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p_\mathbf{\theta}(\mathbf{x}|\mathbf{z})$, as
specified through a deterministic function $g(\cdot; \theta) : \mathbb{R}^m \to
\mathbb{R}^d$. That is
\begin{equation}\label{eqn:p_x_sim}
    p_\mathbf{\theta}(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\mathbf{\theta}) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
\end{equation}
The simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. Importantly, and as commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$, e.g. when specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r$, our goal is the inference of the parameters
of interest $\mathbf{\theta}^*$ that minimize the divergence between $p_r$ and
the modeled data distribution $p_\mathbf{\theta}$ induced by $g(\cdot;
\mathbf{\theta})$ over $\mathbf{z}$. That is,
\begin{equation}
    \mathbf{\theta}^* = \arg \min_\mathbf{\theta} \rho(p_r, p_\mathbf{\theta}),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\mathbf{z}$. More specifically,
a generative model $g(\cdot; \mathbf{\theta})$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $g(\mathbf{z}; \mathbf{\theta})$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to maximally confuse
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r \parallel p_\theta)$ between $p_r$ and $p_\theta$.

@@

Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W}) is a general
optimization technique that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize, VO
is based on the simple fact that
\begin{equation}
    \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c}) \leq \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] = U(\psi),
\end{equation}
where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\mathbf{c}$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\mathbf{c}^* = \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c})$.

@@

\glnote{Naive approach by using VO on ${\cal L}_g$.}
\glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}

>>>>

\section{Problem statement}
\label{sec:problem}

We consider a family of parameterized densities $p_\theta(\mathbf{x})$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in \mathbb{R}^m$ is a latent variable providing an external source
of randomness.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
specified through a deterministic function $g(\cdot; \theta) : \mathbb{R}^m \to
\mathbb{R}^d$. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_z, \mathbf{x} = g(\mathbf{z}; \theta)
\end{equation}
such that the likelihood $p_\theta(\mathbf{x})$ can be rewritten as
\begin{equation}\label{eqn:p_x_sim}
    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
\end{equation}
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r$, our goal is the inference of the parameters
of interest $\theta^*$ that minimize the divergence between $p_r$ and
the modeled data distribution $p_\theta$ induced by $g(\cdot;
\theta)$ over $\mathbf{z}$. That is,
\begin{equation}
    \theta^* = \arg \min_\theta \rho(p_r, p_\theta),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\mathbf{z}$. More specifically,
a generative model $g(\cdot; \theta)$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to maximally confuse
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r \parallel p_\theta)$ between $p_r$ and $p_\theta$.

@@

Following \citep{2012arXiv1212.4507S}, variational optimization (VO) (also known as the search gradient algorithm~\citep{2011arXiv1106.4487W} related to evolution strategies) is a general
optimization technique that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize, VO
is based on the simple fact that
\begin{equation}
    \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c}) \leq \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] = U(\psi),
\end{equation}
where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\mathbf{c}$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\mathbf{c}^* = \min_{\mathbf{c} \in {\cal C}} f(\mathbf{c})$.

@@

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs inherently assume that the generator $g$ is a differentiable function. In
the setting where we are not interested in learning an implicit model but are
rather interested in the inference of parameters of a fixed
non-differentiable simulator (as outlined in Section~\ref{sec:problem}),
gradients $\nabla_\theta g$ either do not exist or cannot be accessed. As a
result, gradients $\nabla_\theta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to perform variational optimization
on ${\cal L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability
of $g$. More specifically, we consider a proposal distribution $q_\psi(\theta)$
over the parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align}
respectively over $\phi$ and $\psi$.
When updating $d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be
obtained by sampling mini-batches of true and generated data, as ordinarily
done in stochastic gradient descent. When updating $g$, gradient estimates
of $\nabla_\psi U_g$ can be derived as described in Eqn.~\ref{eqn:vo-grad},
which gives
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)].
\end{equation}
\glnote{Smarter approach by exploiting the fact that $\partial d / \partial x$ is known exactly.}

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose collection
of generated samples is optimized to approach the real data distribution $p_r$.
